---
title: "Output"
author: "Sam M Sandoval"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

GitHub repository URL: https://github.com/sammsandoval/calypte-anna

Comparing Sensors to Videography for Behavioral Assessment in Anna’s Hummingbirds (Calypte anna)

Bird ethologists often conduct research by scoring ethograms of direct observations or videos of animals either in the field or captivity. However, video scoring can be time consuming, making long-term behavioral assessment difficult. This project aims to develop a method of recording behaviors of Anna’s Hummingbirds (Calypte anna) using affordable small sensors. A cage was constructed consisting of two perches, each attached to a single-point load cell that recorded presence and absence on the perch. As hummingbirds typically do not perch on the floor, the amount of time in flight was determined by the time spent off the perch. Four feeders were equipped with infrared beam sensors to record the time and duration of feeding events. Sensor data was compared with traditional behavioral scoring. Initial results demonstrate a high level of fidelity between sensors and scored video recordings. Certain behaviors could also be associated with specific “signatures” in the scale data, such as preening, suggesting that the perch sensors may be sensitive for measuring activities while perched. This method could reduce the need for human scoring and be applied to many other model species.

This project will create a foundation for the behavioral analyses I will use for my thesis. Fifteen Anna's Hummingbird individuals will be placed under four different photoperiods (daylengths). Food will be provided ad libitum, to prevent any effect of food availability. Scaled perches will monitor changes in activity levels and body mass, while feeders outfitted with sensors will monitor feeding frequency and duration. Before I can begin monitoring these individuals, I need to test if the sensors I have built will gather accurate data. These sensors should be able to record all these activities without any human interaction.

Null Hypothesis: Sensor data will not be as accurate as video data (i.e. there will be a difference between the sensor data and the video data).

Alternative Hypothesis: Sensor data will be as accurate as video data (i.e there will be no difference between sensor data and video data).

Aim 1: Gather behavioral data and prepare it for R studio

Aim 2: Compare video and sensor data using R studio.

The Jaccard/Tanimoto index/coeffiecient statistic will be used to test the similarity of my two datasets and if the difference (or lack thereof) is statistically significant. This index is used for a variety of applications that generate binary data. It compares data by dividing the size of the intersection (the data points that are both the same value) by the size of the union (both data sets together). The resulting value is how similar the datasets are to each other; a large coefficient means they are very similar. In mathematical terms: 

J = Jaccard similarity coefficient
M11 = # of attributes where both sets have a value of 1
M01 = # of attributes where the 1st set has a value of 0 and the 2nd has a value of 1
M10 = # of attributes where the 1st set has a value of 1 and the second has a value of 0

J = M11/(M11 + M01 + M10)

### Load Packages

```{r load-packages, message=FALSE}
#Load packages
library(tidyverse)
library(openintro)
library(ggplot2)
library(devtools)
```
```{r}
#Statistical analysis needs the package "jaccard", but the package "qvalue" 
#should be installed first.

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("qvalue")

install_github("ncchung/jaccard")
library(jaccard)
```

### Load Data

```{r}
#Load in the phidgets sensor scoring data and the video scoring data.

phidgets<- read.csv("phidgets_data_Bird5_Scoring.csv", header = TRUE, stringsAsFactors = TRUE)
str(phidgets)

video <- read.csv("Trial_1_Bird_5_Video_Scoring.csv", header = TRUE, stringsAsFactors = TRUE)
str(video)
```

### Cut Out Unneccessary Data
```{r}
#The video data is not the same length as the phidgets data, so cut it down to
#match.
videosmall <- video[67:1094,]
str(videosmall)

#Focus just on the visitation data for both datasets, since this is the
# behavior we are comparing. Delete the Perch, Hover, and Preen columns

videosmall$Perch <- NULL
videosmall$Preening <- NULL
videosmall$Hover <- NULL

phidgets$Perch <- NULL
phidgets$Preening <- NULL
phidgets$Hover <- NULL
```
### Make Time Data Readable
```{r}
#Tell R that the time data should be treated as time

phidgets$Time <- as.POSIXct(phidgets$Time, format = "%H:%M:%OS")
videosmall$Time <- as.POSIXct(videosmall$Time, format= "%H:%M:%OS")
```

### Visualize Datasets
```{r}
#Create a scatterplot of the phidgets data to see what times a visitation 
#occurred. 

p <- ggplot(data = phidgets, aes(x = Time, y = Visitation)) +
  geom_point() +
  scale_x_datetime(date_breaks = "20 sec", date_labels = "%H:%M:%OS") +
  theme(axis.text.x = element_text(size = 7, angle = 90))
p
```
```{r}

#Create a scatterplot of the video small data to see what times a visitation 
#occurred
v <- ggplot(data = videosmall, aes(x = Time, y = Visitation)) +
  geom_point() +
  scale_x_datetime(date_breaks = "20 sec", date_labels = "%H:%M:%OS") +
  theme(axis.text.x = element_text(size = 7, angle = 90))
v

```

### Statistical Analysis
```{r}
#Statistical Analysis code found at:
#https://github.com/ncchung/jaccard/tree/master
#https://github.com/ncchung/jaccard/blob/master/R/jaccard.test.bootstrap.R

#Rename datasets
x <- videosmall
y <- phidgets

# To compute the exact p-value of similarity between two binary vectors, x and 
#y:

#jaccard.test(x,y,method="exact")$pvalue

# Using a bootstrap method to compare datasets

#jaccard.test.bootstrap <- function(x, y, px = NULL, py = NULL, verbose=TRUE, fix="x", B=1000, seed=NULL) {
#  if(!is.null(123)) set.seed(123)
  # length of fingerprints
#  if(length(x) != length(y)) stop("Length mismatch")
#  m <- length(x)

  # probabilities of ones
#  if(is.null(px) | is.null(py)){
#    px <- mean(x)
#    py <- mean(y)
#  }

#  expectation <- jaccard.ev(x, y, px=px, py=py)
#  j.obs <- jaccard(x, y, center=TRUE, px=px, py=py)
#  degenerate <- FALSE
#  if(px==1 | py==1 | sum(x) == length(x) | sum(y) == length(y)) {
#    warning("One or both input vectors contain only 1's.")
#    degenerate <- TRUE

#  }
#  if(px==0 | py==0 | sum(x) == 0 | sum(y) == 0) {
#    warning("One or both input vectors contain only 0's")
#    degenerate <- TRUE
#  }
#  if(degenerate) {
#    return(list(statistics = 0, pvalue = 1, expectation = expectation))
#  }

#  j.null <- vector("numeric",B)
#  if(verbose) message("Bootstrap Procedures : ")
#  for(i in 1:B) {
#    if(verbose & (i %% floor(B/10)) == 0) {
#      message(paste0(i, " "),appendLF=FALSE)
#    }

#    if(fix == "x") {
#      j.null[i] <- jaccard(x, sample(y, replace=TRUE), center=TRUE)
#    } else if(fix == "y") {
#      j.null[i] <- jaccard(sample(x, replace=TRUE), y, center=TRUE)
#    } else {
#      j.null[i] <- jaccard(sample(x, replace=TRUE), sample(y, replace=TRUE), center=TRUE)
#    }
#  }

#  pvalue <- getp(abs(j.obs), abs(j.null))

#  return(
#    list(
#      statistics = j.obs,
#      statistics.null = j.null,
#      pvalue = pvalue,
#      expectation = expectation)
#  )
#}
```













